---
title: "Processed scRNA-seq & scATAC-seq of mouse gastrulation"
output: 
  html_document:
    toc: true
    toc_depth: 5
    code_folding: hide
    toc_float: true
    code_download: true
    theme: cosmo
    highlight: textmate
---


<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE, 
                      collapse = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")
setwd("/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")

set.seed(1)
```



```{r}
#library(ArchR)
library(knitr)
library(tidyverse)
library(Seurat)
library(SeuratData)
library(scater)
library(zellkonverter)
library(SingleCellExperiment)
library(EnsDb.Mmusculus.v79)
library(Signac)
#library(SeuratDisk)
#library(caret)
#h5disableFileLocking()
```

# Mouse gastrulation days E7.5, E8.0, E8.5, E8.75

We can read in the .h5ad file as a SummarizedCellExperiment

```{r}
rna_gastr_SE <- readH5AD("jupyter_notebooks/anndata_rna.h5ad")
atac_gastr_SE <- readH5AD("jupyter_notebooks/anndata_atac_peak_matrix.h5ad")

print(paste0("The RNA data has dimensions ", paste(dim(rna_gastr_SE), collapse = ", ")))
print(paste0("The ATAC data has dimensions ", paste(dim(atac_gastr_SE), collapse = ", ")))
```


We can convert a Seurat object to a SummarizedCellExperiment

```
SCE <- as.SingleCellExperiment(gastr_seurat)
```

We can convert SummarizedCellExperiment to a Seurat object. The SummarizedExperiment
contains the raw counts of the gene expression. Additionally we can add the
metadata information to the Seurat objects.

```{r, results = "asis"}
rna_seurat <- as.Seurat(rna_gastr_SE, counts = "X", data = "X")
atac_seurat <- as.Seurat(atac_gastr_SE, counts = "X", data = "X")

# Lets add the metadata to the Seurat object
rna_seurat <- AddMetaData(rna_seurat, as.data.frame(colData(rna_gastr_SE)))
atac_seurat <- AddMetaData(atac_seurat, as.data.frame(colData(atac_gastr_SE)))
rna_seurat <- AddMetaData(rna_seurat, atac_seurat@meta.data %>% dplyr::select(BlacklistRatio:FRIP))


rna_seurat@meta.data %>% head %>% knitr::kable()
atac_seurat@meta.data %>% head %>% knitr::kable()
```




# QC


### Empty Droplets

```{r}

tibble(sample = rna_seurat@meta.data$sample, umi_per_cell = Matrix::colSums(rna_seurat@assays$originalexp @counts)) %>% 
  arrange(sample, desc(umi_per_cell)) %>% 
  group_by(sample) %>% 
  mutate(idx = seq_along(sample)) %>% 
  mutate(cum_umi_per_cell = cumsum(umi_per_cell)) %>% 
  ggplot() +
  geom_line(aes(x = idx, y = cum_umi_per_cell)) +
  facet_wrap(~sample, scales = "fixed") +
  scale_y_log10() + scale_x_log10() +
  ylab("cumulative UMI count") +
  xlab("index")


```



From the plots below it seems that all cells with percentage of mitochondrial 
genes above 50% were removed.


```{r}
#rename metadata

variables <- c("nFeature_originalexp", "nCount_originalexp", "mitochondrial_percent_RNA", "ribosomal_percent_RNA")
map(variables, function(n){
  df <- rna_seurat@meta.data
  ggplot(df) +
    geom_boxplot(aes(x = df %>% pull("sample"), y = df %>% pull(n),
                     fill = df %>% pull("sample")), alpha = .1) +
    geom_violin(aes(x = df %>% pull("sample"), y = df %>% pull(n),
                     fill = df %>% pull("sample")), alpha = .5) +
    xlab("sample") +
    ylab(paste0(n)) +
    labs(title = paste0(n)) +
    guides(fill=guide_legend(title="sample"))
})

```

## How good are the annotations?

There are two different methods for mapping listed in the metadata:

* Seurat
* MNN

```{r}
atac_seurat@meta.data %>% group_by(celltype.mapped_seurat) %>% 
  summarise(n = n()) %>% knitr::kable(caption = "Number of celltypes using Seurat")

atac_seurat@meta.data %>% group_by(celltype.mapped_mnn) %>% 
  summarise(n = n()) %>% knitr::kable(caption = "Number of celltypes using MNN")
```



The annotations are similar across the different samples. However, there are
some differences in the confidence with which cells can me mapped for different
celltypes. For example, cardiomyocytes, erythroids and extraembryonic ectoderm
and endoderm (but not mesoderm), mesenchyme and parietal endoderm can be mapped
with very high confidence. 

```{r, fig.height=8, fig.width=8}
atac_seurat@meta.data %>% 
  ggplot() + 
  geom_boxplot(aes(x = celltype.mapped_seurat, y =  celltype.score_seurat, fill = celltype.mapped_seurat)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.8, hjust=1)) + NoLegend()


atac_seurat@meta.data %>% 
  ggplot() + 
  geom_boxplot(aes(x = celltype.mapped_mnn, y =  celltype.score_mnn, fill = celltype.mapped_mnn)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.8, hjust=1)) + NoLegend()
```


## Normalize & Scale

```{r}
rna_seurat <- rna_seurat %>% 
  NormalizeData(verbose = FALSE) %>% 
  ScaleData(verbose = FALSE) %>%
  FindVariableFeatures(verbose = FALSE)


#gastr_seurat@assays$RNA@data[0:10, 0:10]
```


### PCA

I will proceed with 15 PCs.

```{r}
rna_seurat <- RunPCA(rna_seurat, features = VariableFeatures(rna_seurat),
             verbose = FALSE)

ElbowPlot(rna_seurat)
```

```{r}
pca_plots <- comprehenr::to_list(for (i in 1:15)
  DimPlot(rna_seurat, reduction = "pca", dims = i:(i+1)) +
    theme())

#pca_plots
#gridExtra::grid.arrange(unlist(pca_plots), ncol = 3, nrow = 5)

```

## Clustering

```{r}
rna_seurat <- FindNeighbors(rna_seurat, verbose = FALSE)
rna_seurat <- FindClusters(rna_seurat, verbose = FALSE, resolution = 0.9)
rna_seurat <- RunUMAP(rna_seurat, verbose = FALSE, dims = 1:15)
```


# Visualization

```{r}
colPalette_celltypes = c('#532C8A',
 '#c19f70',
 '#f9decf',
 '#c9a997',
 '#B51D8D',
 '#3F84AA',
 '#9e6762',
 '#354E23',
 '#F397C0',
 '#ff891c',
 '#635547',
 '#C72228',
 '#f79083',
 '#EF4E22',
 '#989898',
 '#7F6874',
 '#8870ad',
 '#647a4f',
 '#EF5A9D',
 '#FBBE92',
 '#139992',
 '#cc7818',
 '#DFCDE4',
 '#8EC792',
 '#C594BF',
 '#C3C388',
 '#0F4A9C',
 '#FACB12',
 '#8DB5CE',
 '#1A1A1A',
 '#C9EBFB',
 '#DABE99',
 '#65A83E',
 '#005579',
 '#CDE088',
 '#f7f79e',
 '#F6BFCB')
```


There are two possibilities to proceed with celltype annotations:

* Seurat
* MNN

```{r}
celltypes <- (atac_seurat@meta.data %>% group_by(celltype.mapped_seurat) %>% 
  summarise(n = n()))$celltype.mapped_seurat

col <- setNames(colPalette_celltypes, celltypes)
```


```{r, fig.width=15, fig.height = 10}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", label = TRUE, repel = TRUE, cols = col) +
  NoLegend()
```


```{r, fig.width=15, fig.height = 10}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", cols = col) #, label = TRUE, repel = TRUE) +
```

```{r}
DimPlot(rna_seurat, reduction = "umap", pt.size = .1, 
        group.by = "seurat_clusters", label = TRUE) +
  NoLegend()
```

```{r, fig.width=15, fig.height=10}
p1 <- DimPlot(rna_seurat, reduction = "umap", pt.size = .1, group.by = "sample")

p2 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1,
            features = "nCount_RNA") +
    scale_color_viridis_c() 
p3 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1, features = "mitochondrial_percent_RNA") +
    scale_color_viridis_c() 

p4 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1, features = "nFeature_RNA") +
    scale_color_viridis_c() 

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```


```{r, fig.width=15, fig.height=20}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", split.by = "orig.ident", ncol = 1, cols = col) +
  labs(title = "Celltypes at different time points")
```



```#{r, fig.width = 12, fig.height=8}
rna_seurat@meta.data %>%
  group_by(orig.ident, celltype.mapped) %>% 
  summarise(Total = n()) %>% 
  mutate(freq = Total/sum(Total)) %>% 
  ggplot(aes(x = celltype.mapped, y = freq, fill = orig.ident)) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust=1)) +
  facet_wrap(~orig.ident, ncol = 1)
```

We can see that nascent mesoderm, epiblast and primitive streak are only present 
at E75. Also, extraembryonice endoderm and ectoderm are highest and decrease 
with progressing gastrulation. Forebrain/Midbrain/Hindbrain, neural crest and 
neuromesodermal progenitor (NMP) cells are not present at 
E7.5, but emerge at E8.0 and are present at even higher percentage at E8.5. The 
same is true for Allantois, erythroids (produce red blood cells, remain in bone
marrow)and cardiomyocytes.

(Pijuan_Sala et.al, A single-cell molecular map of mouse
gastrulation and early organogenesis, 2019, Nature)

```{r, fig.width=10}
# plot the frequency of each cell type at each embryonic stage
# all frequencies for one embryonic stage would add up to 0
rna_seurat@meta.data %>%
  group_by(orig.ident, celltype.mapped_seurat) %>% 
  summarise(Total = n()) %>% 
  mutate(freq = Total/sum(Total)) %>% 
  ggplot(aes(x = celltype.mapped_seurat, y = freq, fill = orig.ident)) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust=1)) 
  
```



```{r}
subset(rna_seurat, subset = orig.ident == "E7.5")
```

### Processing Data at each time step separately

```{r}
stages <- c("E7.5", "E8.0", "E8.5", "E8.75")
seurat_objects <- map(stages, function(n){
  rna_seurat <- subset(rna_seurat, subset = orig.ident == n)
  rna_seurat <- rna_seurat  %>% 
    NormalizeData(verbose = FALSE) %>% 
    ScaleData(verbose = FALSE) %>%
    FindVariableFeatures(verbose = FALSE) 
  rna_seurat <- rna_seurat %>%  RunPCA(features = VariableFeatures(rna_seurat), verbose = FALSE) %>% 
    FindNeighbors(verbose = FALSE) %>% 
    FindClusters(verbose = FALSE, resolution = .9) %>% 
    RunUMAP(verbose = TRUE, dims = 1:15)
  list(name = n, object = rna_seurat)
    
})


e75 <- seurat_objects[[1]]$object
e80 <- seurat_objects[[2]]$object
e85 <- seurat_objects[[3]]$object
e785 <- seurat_objects[[4]]$object
```


```#{r}
umaps <- comprehenr::to_list(for (i in 1:4)
  DimPlot(rna_seurat, reduction = "pca", dims = i:(i+1)) +
    theme()) 
```


```{r, fig.width=15, fig.height=20}
plots <- map(seq.int(1,4), function(n){
  object <- seurat_objects[[n]]$object
  p1 <- DimPlot(object, reduction = "umap", group.by = "celltype.mapped_seurat", 
                cols = col, size = .9) +
    labs(title = paste0(seurat_objects[[n]]$name, " celltypes"))
  list(plot = p1)
})

gridExtra::grid.arrange(plots[[1]]$p, plots[[2]]$p, plots[[3]]$p,plots[[4]]$p, ncol = 1)
```



# scATAC-seq data



```
# genome annotations from EnsDb
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)

# change to UCSC style 
seqlevelsStyle(annotations) <- "UCSC"

# add gene information to the seurat object
Annotation(atac_seurat) <- annotations
```


# QC ATAC

```{r}
atac_seurat@meta.data %>%
  ggplot() +
  geom_density2d_filled(aes(x=log10(nFrags_atac ), y=TSSEnrichment_atac), bins=20) +
  geom_hline(yintercept = 5, color="green", linetype="dashed") +
  geom_vline(xintercept = 3, color="green", linetype="dashed") +
  #geom_xsidedensity(aes(x=log10(pre_filter_meta$nFrags))) +
  #geom_ysidedensity(aes(y = pre_filter_meta$TSSEnrichment)) +
  facet_wrap(~sample) +
  theme(legend.position = "none") +
  labs(x = "Log10 Unique Fragments", y = "TSS Enrichment Score")
```



```{r, fig.width=10}
p1 <- atac_seurat@meta.data %>% 
  ggplot() +
  ggridges::geom_density_ridges(aes(x = TSSEnrichment, y = Sample, fill = Sample),
                                alpha = .6)

p2 <- atac_seurat@meta.data %>% 
  ggplot() +
  geom_violin(aes(x = Sample, y = TSSEnrichment, fill = Sample), alpha = 0.6) +
  geom_boxplot(aes(x = Sample, y = TSSEnrichment,fill = Sample), alpha = 0.1) + 
  theme(legend.position = "none") +
  labs(title = "TSS Enrichment")
cowplot::plot_grid(p2, p1, ncol = 2)

```

```{r, fig.width=10}
p1 <- atac_seurat@meta.data %>% 
  ggplot() +
  ggridges::geom_density_ridges(aes(x = nFrags_atac , y = Sample, fill = Sample),
                                alpha = .6)

p2 <- atac_seurat@meta.data %>% 
  ggplot() +
  geom_violin(aes(x = Sample, y = nFrags_atac , fill = Sample), alpha = 0.6) +
  geom_boxplot(aes(x = Sample, y = nFrags_atac ,fill = Sample), alpha = 0.1) + 
  theme(legend.position = "none") +
  labs(title = "TSS Enrichment")
cowplot::plot_grid(p2, p1, ncol = 2)
```

# Normalization & Dimensionality reduction ATAC

**Latent Semantic Indexing**

* TF-IDF normalization
  + normalizes across cells (sequencing depth)
  + across peak (higher values to more rare peaks)
* Select Top features
* Run SVD on selected features

```{r}
atac_seurat <- RunTFIDF(atac_seurat)
atac_seurat <- FindTopFeatures(atac_seurat)
atac_seurat <- RunSVD(atac_seurat)
```
The first LSI component often captures sequencing depth. We will
therefore remove it from downstream analysis. The correlation between sequencing
depth and each LSI component is shown in the plot below.

```{r}
DepthCor(atac_seurat)
```

# Visualization

```{r}
atac_seurat <- RunUMAP(atac_seurat, reduction = "lsi", dims = 2:30)
aatac_seurat <- FindNeighbors(atac_seurat, reduction = "lsi", dims = 2:30)

# for Clsutering instead of Louvian SLM algorithm is used
#atac_seurat <- FindClusters(atac_seurat, verbose = FALSE, algorithm = 3) 
```



```{r, fig.width = 15, fig.height=10}

DimPlot(atac_seurat, group.by = "celltype.mapped_seurat", reduction = "umap", pt.size = 1, cols = col)# +
  labs(title = "scATAC-seq Celltype")
  
DimPlot(atac_seurat, group.by = "sample", reduction = "umap", pt.size = 1, cols = col)# +
  labs(title = "scATAC-seq Celltype")
```

```{r, fig.width=15, fig.height=10}
DimPlot(rna_seurat , group.by = "celltype.mapped_seurat", pt.size = 1, cols = col,
        reduction = "umap") +
  labs(title = "scRNA-seq Celltype")

```





```{r, fig.width=10, fig.height=8}
atac_seurat@meta.data %>% 
  ggplot(aes(x = log10(nCount_originalexp ), y = log10(nFrags_atac))) +
  geom_point(alpha = .2, size = .2) +
  ggside::geom_xsidedensity() +
  ggside::geom_ysidedensity() +
  facet_wrap(~sample) +
  labs(x = "Log10 Counts", y = "log10 Unique Fragments")
```


```{r, fig.width=10, fig.height=10}
atac_seurat@meta.data %>% 
  ggplot(aes(x = log10(nCount_originalexp), y = log10(TSSEnrichment))) +
  geom_point(size = .2, alpha = .2) +
  ggside::geom_xsidedensity() +
  ggside::geom_ysidedensity() +
  facet_wrap(~sample)
```


```{r}
atac_seurat@meta.data %>% 
  ggplot() +
  geom_histogram(aes(x = PromoterRatio_atac)) 

atac_seurat@meta.data %>% 
  ggplot() +
  geom_histogram(aes(x = NucleosomeRatio_atac)) 
```




# Integrating RNA & ATAC

We will remove all cells from the scRNA-seq dataset which do not pass the 
scATAC-seq QC and vice versa.

```{r}
#rna_seurat_filt <- subset(gastr_seurat, pass_atacQC == "True")
# atac_seurat_filt <- subset(atac_seurat, pass_rnaQC == TRUE)

```

