---
title: "Processed scRNA-seq & scATAC-seq of mouse gastrulation"
output: 
  html_document:
    toc: true
    toc_depth: 5
    code_folding: hide
    toc_float: true
    code_download: true
    theme: cosmo
    highlight: textmate
---


<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE, 
                      collapse = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")
setwd("/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")

set.seed(1)
```



```{r}
suppressPackageStartupMessages({
library(knitr)
library(tidyverse)
library(Seurat)
library(SeuratData)
library(scater)
library(zellkonverter)
library(SingleCellExperiment)
library(EnsDb.Mmusculus.v79)
library(Signac)
})
```

# Mouse gastrulation days E7.5, E8.0, E8.5, E8.75

The dataset consists of multiome data (scATAC-seq & scRNA-seq) from cells at the 
four mentioned timepoints.

We can read in the .h5ad file as a SummarizedCellExperiment

```{r}
rna_gastr_SE <- readH5AD("jupyter_notebooks/anndata_rna.h5ad")
atac_gastr_SE <- readH5AD("jupyter_notebooks/anndata_atac_peak_matrix.h5ad")

print(paste0("The RNA data has dimensions ", paste(dim(rna_gastr_SE), collapse = ", ")))
print(paste0("The ATAC data has dimensions ", paste(dim(atac_gastr_SE), collapse = ", ")))
```


We can convert SummarizedCellExperiment to a Seurat object. The SummarizedExperiment
contains the raw counts of the gene expression. Additionally we can add the
metadata information to the Seurat objects.

```{r, results = "asis"}
rna_seurat <- as.Seurat(rna_gastr_SE, counts = "X", data = "X")
atac_seurat <- as.Seurat(atac_gastr_SE, counts = "X", data = "X")

# Lets add the metadata to the Seurat object
rna_seurat <- AddMetaData(rna_seurat, as.data.frame(colData(rna_gastr_SE)))
atac_seurat <- AddMetaData(atac_seurat, as.data.frame(colData(atac_gastr_SE)))
rna_seurat <- AddMetaData(rna_seurat, atac_seurat@meta.data %>% dplyr::select(BlacklistRatio:FRIP))


rna_seurat@meta.data %>% head %>% knitr::kable()
atac_seurat@meta.data %>% head %>% knitr::kable()
```


# scRNA-seq

## QC


### Empty Droplets

```{r}

tibble(sample = rna_seurat@meta.data$sample, umi_per_cell = Matrix::colSums(rna_seurat@assays$originalexp @counts)) %>% 
  arrange(sample, desc(umi_per_cell)) %>% 
  group_by(sample) %>% 
  mutate(idx = seq_along(sample)) %>% 
  mutate(cum_umi_per_cell = cumsum(umi_per_cell)) %>% 
  ggplot() +
  geom_line(aes(x = idx, y = cum_umi_per_cell)) +
  facet_wrap(~sample, scales = "fixed") +
  scale_y_log10() + scale_x_log10() +
  ylab("cumulative UMI count") +
  xlab("index")


```



From the plots below it seems that all cells with percentage of mitochondrial 
genes above 40% were removed. The early time points E7.5rep1/2 have the highest
percentage of mitochondrial genes. Especially E7.5 rep2 seems to have a higher 
percentage of mitochondrial genes. Conversely, the same replicates from E7.5 
have a lower number of features and counts. This means that the E7.5 rep2 and to
a lesser extent E7.5 rep1 are of lower quality compared to the other samples. E8.75 
rep1/rep2 seem to have the highest quality. 


```{r, fig.width=15, 10}
#rename metadata

variables <- c("nFeature_originalexp", "nCount_originalexp", "mitochondrial_percent_RNA", "ribosomal_percent_RNA")
plots <- map(variables, function(n){
  df <- rna_seurat@meta.data
  ggplot(df) +
    geom_boxplot(aes(x = df %>% pull("sample"), y = df %>% pull(n),
                     fill = df %>% pull("sample")), alpha = .1) +
    geom_violin(aes(x = df %>% pull("sample"), y = df %>% pull(n),
                     fill = df %>% pull("sample")), alpha = .5) +
    xlab("sample") +
    ylab(paste0(n)) +
    labs(title = paste0(n)) +
    guides(fill=guide_legend(title="sample"))
})

gridExtra::grid.arrange(grobs = plots, ncol = 2)

```

## How good are the annotations?

There are two different methods for mapping listed in the metadata:

* Seurat
* MNN

```{r}
atac_seurat@meta.data %>% group_by(celltype.mapped_seurat) %>% 
  summarise(n = n()) %>% knitr::kable(caption = "Number of celltypes using Seurat")

atac_seurat@meta.data %>% group_by(celltype.mapped_mnn) %>% 
  summarise(n = n()) %>% knitr::kable(caption = "Number of celltypes using MNN")
```



The annotations are similar across the different samples. However, there are
some differences in the confidence with which cells can me mapped for different
celltypes. For example, cardiomyocytes, erythroids and extraembryonic ectoderm
and endoderm (but not mesoderm), mesenchyme and parietal endoderm can be mapped
with very high confidence. 

```{r, fig.height=8, fig.width=8}
atac_seurat@meta.data %>% 
  ggplot() + 
  geom_boxplot(aes(x = celltype.mapped_seurat, y =  celltype.score_seurat, fill = celltype.mapped_seurat)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.8, hjust=1)) + NoLegend()


atac_seurat@meta.data %>% 
  ggplot() + 
  geom_boxplot(aes(x = celltype.mapped_mnn, y =  celltype.score_mnn, fill = celltype.mapped_mnn)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.8, hjust=1)) + NoLegend()
```


## Normalize & Scale

```{r}
rna_seurat <- rna_seurat %>% 
  NormalizeData(verbose = FALSE) %>% 
  ScaleData(verbose = FALSE) %>%
  FindVariableFeatures(verbose = FALSE)


#gastr_seurat@assays$RNA@data[0:10, 0:10]
```


### PCA

I will proceed with 15 PCs.

```{r}
rna_seurat <- RunPCA(rna_seurat, features = VariableFeatures(rna_seurat),
             verbose = FALSE)

ElbowPlot(rna_seurat)
```

```{r}
pca_plots <- comprehenr::to_list(for (i in 1:15)
  DimPlot(rna_seurat, reduction = "pca", dims = i:(i+1)) +
    theme())

#pca_plots
#gridExtra::grid.arrange(unlist(pca_plots), ncol = 3, nrow = 5)

```

## Clustering

Trying a different resolution for the clustering to see if the Mesenchyme will 
still separate.

```{r}
rna_seurat <- FindNeighbors(rna_seurat, verbose = FALSE)
rna_seurat <- FindClusters(rna_seurat, verbose = FALSE, resolution = 0.6)
rna_seurat <- RunUMAP(rna_seurat, verbose = FALSE, dims = 1:15)
```


## Visualization

```{r}
colPalette_celltypes = c('#532C8A',
 '#c19f70',
 '#f9decf',
 '#c9a997',
 '#B51D8D',
 '#3F84AA',
 '#9e6762',
 '#354E23',
 '#F397C0',
 '#ff891c',
 '#635547',
 '#C72228',
 '#f79083',
 '#EF4E22',
 '#989898',
 '#7F6874',
 '#8870ad',
 '#647a4f',
 '#EF5A9D',
 '#FBBE92',
 '#139992',
 '#cc7818',
 '#DFCDE4',
 '#8EC792',
 '#C594BF',
 '#C3C388',
 '#0F4A9C',
 '#FACB12',
 '#8DB5CE',
 '#1A1A1A',
 '#C9EBFB',
 '#DABE99',
 '#65A83E',
 '#005579',
 '#CDE088',
 '#f7f79e',
 '#F6BFCB')
```


There are two possibilities to proceed with celltype annotations:

* Seurat (this is the one I used)
* MNN

In the plot below you can see that the Mesenchyme separates into two clusters, 
even though the cells belong to the same celltype. Therefore I visualized the 
Mesenchyme cells according to their respective time point and replicate below. It 
becomes evident that the Mesenchyme separate into E7.5 and E8.0, E8.5, E8.75. Therefore,
the separation is probably a biological effect corresponding to different signatures 
at different time points. However, we should keep in mind that E7.5 was also the
time point with lowest quality of cells. 

There is one cluster which contains a very heterogeneous population of cells,
namely mixed & nascent mesoderm, caudal & rostral neuroectoderm, primitive streak, 
caudal epiblast and epiblast. As we will see later on, these are primarily early celltypes
found at E7.5, which are not present at later timepoints anymore.

```{r}
celltypes <- (atac_seurat@meta.data %>% group_by(celltype.mapped_seurat) %>% 
  summarise(n = n()))$celltype.mapped_seurat

col <- setNames(colPalette_celltypes, celltypes)
```


```{r, fig.width=15, fig.height = 10}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", label = TRUE, repel = TRUE, cols = col) +
  NoLegend()
```

```{r}
DimPlot(subset(rna_seurat, celltype.mapped_seurat == "Mesenchyme"), 
        group.by = "sample")
DimPlot(subset(rna_seurat, celltype.mapped_seurat == "Mesenchyme"), 
        split.by  = "sample")
```

Plot with color legend.

```{r, fig.width=15, fig.height = 10}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", cols = col) #, label = TRUE, repel = TRUE) +
```

Clustering by Seurat with resolution = 0.6 yields 17 distinct clusters, not enough
to differentiate the large number of different celltypes in this dataset.

```{r}
DimPlot(rna_seurat, reduction = "umap", pt.size = .1, 
        group.by = "seurat_clusters", label = TRUE) +
  NoLegend()
```

Below, you can see how the earlier time points separate from the later time 
points. The number of counts is very homogenous, while the mitochondrial RNA 
percentage is higher and the number of features lower for E7.5 as already pointed out.

```{r, fig.width=15, fig.height=10}
p1 <- DimPlot(rna_seurat, reduction = "umap", pt.size = .1, group.by = "sample")

p2 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1,
            features = "nCount_RNA") +
    scale_color_viridis_c() 
p3 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1, features = "mitochondrial_percent_RNA") +
    scale_color_viridis_c() 

p4 <- FeaturePlot(rna_seurat, reduction = "umap", pt.size = .1, features = "nFeature_RNA") +
    scale_color_viridis_c() 

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```


## Distribution of celltypes at different time points


Visualizing the cells at each timepoint in separate UMAPs shows how the
distribution of celltypes changes as gastrulation proceeds. The heterogenous 
cell cluster present at E7.5 disappears at later time points. We can also see how
no Erythroids are present at E7.5,
Erythroid1 (and to lesser extend Erythroid2 and 3) appear at E8.0, while Eryhtroid3 
are found to a higher extent
at E8.75, which indicates that these cells correspond to different developmental 
stages. 

```{r, fig.width=15, fig.height=20}
DimPlot(rna_seurat, reduction = "umap", pt.size = 1, 
        group.by = "celltype.mapped_seurat", split.by = "orig.ident", ncol = 1, cols = col) +
  labs(title = "Celltypes at different time points")
```


We can see that nascent mesoderm, epiblast and primitive streak are only present 
at E75. Also, extraembryonice endoderm and ectoderm are highest and decrease 
with progressing gastrulation. Forebrain/Midbrain/Hindbrain, neural crest and 
neuromesodermal progenitor (NMP) cells are not present at 
E7.5, but emerge at E8.0 and are present at even higher percentage at E8.5. Also,
the endothelium appears only at E8.0. The 
same is true for Allantois, erythroids (produce red blood cells, remain in bone
marrow)and cardiomyocytes.  Conversely, the number
of extraembryonic ectoderm cells and extraembryonic endoderm cells become less 
at each time step. 

(Pijuan_Sala et.al, A single-cell molecular map of mouse
gastrulation and early organogenesis, 2019, Nature)

```{r, fig.width=10}
# plot the frequency of each cell type at each embryonic stage
# all frequencies for one embryonic stage would add up to 0
rna_seurat@meta.data %>%
  group_by(orig.ident, celltype.mapped_seurat) %>% 
  summarise(Total = n()) %>% 
  mutate(freq = Total/sum(Total)) %>% 
  ggplot(aes(x = celltype.mapped_seurat, y = freq, fill = orig.ident)) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.8, hjust=1)) 
  
```


## Processing Data at each time step separately

When doing the preprocessing steps and dimensionality reduction for each 
time point separately, we can create more clearly separated clusters for the
later time points, however, for E7.5 the mixed cluster of epiblast, mixed & nascent
mesoderm, primitive streak and PGC remains a heterogeneous cluster. 

```{r}
stages <- c("E7.5", "E8.0", "E8.5", "E8.75")
seurat_objects <- map(stages, function(n){
  rna_seurat <- subset(rna_seurat, subset = orig.ident == n)
  rna_seurat <- rna_seurat  %>% 
    NormalizeData(verbose = FALSE) %>% 
    ScaleData(verbose = FALSE) %>%
    FindVariableFeatures(verbose = FALSE) 
  rna_seurat <- rna_seurat %>%  RunPCA(features = VariableFeatures(rna_seurat), verbose = FALSE) %>% 
    FindNeighbors(verbose = FALSE) %>% 
    FindClusters(verbose = FALSE, resolution = .9) %>% 
    RunUMAP(verbose = TRUE, dims = 1:15)
  list(name = n, object = rna_seurat)
    
})


e75 <- seurat_objects[[1]]$object
e80 <- seurat_objects[[2]]$object
e85 <- seurat_objects[[3]]$object
e785 <- seurat_objects[[4]]$object
```



```{r, fig.width=15, fig.height=20}
plots <- map(seq.int(1,4), function(n){
  object <- seurat_objects[[n]]$object
  p1 <- DimPlot(object, reduction = "umap", group.by = "celltype.mapped_seurat", 
                cols = col, size = .9) +
    labs(title = paste0(seurat_objects[[n]]$name, " celltypes"))
  list(plot = p1)
})

gridExtra::grid.arrange(plots[[1]]$p, plots[[2]]$p, plots[[3]]$p,plots[[4]]$p, ncol = 1)
```



# scATAC-seq data


## QC ATAC

```{r}
atac_seurat@meta.data %>%
  ggplot() +
  geom_density2d_filled(aes(x=log10(nFrags_atac ), y=TSSEnrichment_atac), bins=20) +
  geom_hline(yintercept = 5, color="green", linetype="dashed") +
  geom_vline(xintercept = 3, color="green", linetype="dashed") +
  #geom_xsidedensity(aes(x=log10(pre_filter_meta$nFrags))) +
  #geom_ysidedensity(aes(y = pre_filter_meta$TSSEnrichment)) +
  facet_wrap(~sample) +
  theme(legend.position = "none") +
  labs(x = "Log10 Unique Fragments", y = "TSS Enrichment Score")
```

The scATAC-seq shows that E7.5 seems to be of lower quality, which we have already
observed in the scRNA-seq. Potentially, the cells where simply in a worse 
condition when sequenced.

```{r, fig.width=10}
p1 <- atac_seurat@meta.data %>% 
  ggplot() +
  ggridges::geom_density_ridges(aes(x = TSSEnrichment, y = Sample, fill = Sample),
                                alpha = .6)

p2 <- atac_seurat@meta.data %>% 
  ggplot() +
  geom_violin(aes(x = Sample, y = TSSEnrichment, fill = Sample), alpha = 0.6) +
  geom_boxplot(aes(x = Sample, y = TSSEnrichment,fill = Sample), alpha = 0.1) + 
  theme(legend.position = "none") +
  labs(title = "TSS Enrichment")
cowplot::plot_grid(p2, p1, ncol = 2)

```

```{r, fig.width=10}
p1 <- atac_seurat@meta.data %>% 
  ggplot() +
  ggridges::geom_density_ridges(aes(x = nFrags_atac , y = Sample, fill = Sample),
                                alpha = .6)

p2 <- atac_seurat@meta.data %>% 
  ggplot() +
  geom_violin(aes(x = Sample, y = nFrags_atac , fill = Sample), alpha = 0.6) +
  geom_boxplot(aes(x = Sample, y = nFrags_atac ,fill = Sample), alpha = 0.1) + 
  theme(legend.position = "none") +
  labs(title = "TSS Enrichment")
cowplot::plot_grid(p2, p1, ncol = 2)
```

## Normalization & Dimensionality reduction ATAC

**Latent Semantic Indexing**

1. TF-IDF normalization
  + normalize across cells (sequencing depth)
  + normalize across peak (higher values to more rare peaks)
2. Select Top features
3.  Run SVD on selected features

```{r}
atac_seurat <- RunTFIDF(atac_seurat)
atac_seurat <- FindTopFeatures(atac_seurat)
atac_seurat <- RunSVD(atac_seurat)
```
The first LSI component often captures sequencing depth. We will
therefore remove it from downstream analysis. The correlation between sequencing
depth and each LSI component is shown in the plot below.

```{r}
DepthCor(atac_seurat)
```

## Visualization

```{r}
atac_seurat <- RunUMAP(atac_seurat, reduction = "lsi", dims = 2:30, verbose = FALSE)
aatac_seurat <- FindNeighbors(atac_seurat, reduction = "lsi", dims = 2:30, verbose = FALSE)

# for Clsutering instead of Louvian SLM algorithm is used
#atac_seurat <- FindClusters(atac_seurat, verbose = FALSE, algorithm = 3) 
```



```{r, fig.width = 15, fig.height=10}

DimPlot(atac_seurat, group.by = "celltype.mapped_seurat", reduction = "umap", pt.size = 1, cols = col) +
  labs(title = "scATAC-seq Celltype")
  
DimPlot(atac_seurat, group.by = "sample", reduction = "umap", pt.size = 1, cols = col) +
  labs(title = "scATAC-seq Celltype")
```

```{r, fig.width=15, fig.height=10}
DimPlot(rna_seurat , group.by = "celltype.mapped_seurat", pt.size = 1, cols = col,
        reduction = "umap") +
  labs(title = "scRNA-seq Celltype")

```


Lets have a look at whether the number of fragments is correlated to the 
number of counts in each celltype. This seems to be the case.


```{r, fig.width=10, fig.height=8}
atac_seurat@meta.data %>% 
  ggplot(aes(x = log10(nCount_originalexp ), y = log10(nFrags_atac))) +
  geom_point(alpha = .2, size = .2) +
  ggside::geom_xsidedensity() +
  ggside::geom_ysidedensity() +
  facet_wrap(~sample) +
  labs(x = "Log10 Counts", y = "log10 Unique Fragments")
```


```{r, fig.width=10, fig.height=10}
atac_seurat@meta.data %>% 
  ggplot(aes(x = log10(nCount_originalexp), y = log10(TSSEnrichment))) +
  geom_point(size = .2, alpha = .2) +
  ggside::geom_xsidedensity() +
  ggside::geom_ysidedensity() +
  facet_wrap(~sample)
```


```{r}
atac_seurat@meta.data %>% 
  ggplot() +
  geom_histogram(aes(x = PromoterRatio_atac)) 

atac_seurat@meta.data %>% 
  ggplot() +
  geom_histogram(aes(x = NucleosomeRatio_atac)) 
```



```#{r}
saveRDS(atac_seurat, "atac_Seurat_object")
saveRDS(rna_seurat, "rna_Seurat_object")
```


