---
title: "Peak2Gene links and gene activity score"
output: 
  html_document:
    toc: true
    toc_depth: 5
    code_folding: hide
    toc_float: true
    code_download: true
    theme: cosmo
    highlight: textmate
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, autodep = TRUE, 
                      collapse = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = "/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")
setwd("/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/")

set.seed(1)
```

  
```{r}
## Load libraries


suppressPackageStartupMessages({
library(ArchR)
library(rhdf5)
library(tidyverse)
library(reticulate)
library(zellkonverter)
library(Matrix)
library(dichromat)
#library(caret)
h5disableFileLocking()})
```


```{r}
proj <- loadArchRProject("12_Ricards_peaks_ChromVar/")


pos_peak_mat <- readH5AD("/omics/groups/OE0533/internal/katharina/scDoRI/gastrulation_data/positive_p2g_link_matrix.h5ad")
pos_peak_mat <- assays(pos_peak_mat)[[1]]
```

```{r}
proj <- addPeak2GeneLinks(  ArchRProj = proj,
  reducedDims  = "atac_LSI_100000",
  useMatrix = "GeneExpressionMatrix",
  maxDist = 400000000,
  verbose = FALSE, 
  )
```

```{r}
p2g <- getPeak2GeneLinks(
  ArchRProj = proj,
  corCutOff = -1,
  resolution = 1,
  FDRCutOff = 1e-04,
  varCutOffATAC = .25,
  varCutOffRNA = .25, 
  returnLoops = FALSE
)
```

## Add min value so we get only positive values

To shift the entire dataset to a positive range, we add the absolute minimum 
value to all correlation values in our dataset. It might make
sense to do this on a gene or peak level. 

```{r}
#p2g_filt <- p2g %>% as.data.frame() %>% filter(Correlation >= 0)
p2g_filt <- p2g

p2g_filt <- p2g_filt %>% as.data.frame() %>%  mutate(pos_values = Correlation + abs(min(Correlation)) + 1) 
```



```{r}
print(paste0("Number of peak2gene links found: ", dim(p2g)))
print(paste0("Total number of peaks called in the dataset: ", length(metadata(p2g)$peakSet)))
print(paste0("For each of the peaks we have metadata containing information on peakType, nearestGene, disttoGeneStart,etc..."))

```


First we will try not to binarize the matrix

```{r}
peaks <- getMatrixFromProject(proj, useMatrix = "PeakMatrix", binarize = FALSE)

# get matrix
peak_mat <- assays(peaks)[[1]]

peakset <- as_tibble(getPeakSet(proj))
peakset <- peakset %>% unite(chr_idx, seqnames, idx, sep = "_", remove = FALSE)
# use index to asssign rownames to the matrix
rownames(peak_mat) <- peakset$chr_idx
```






```#{r}
# extract gene expression SE from the project
gene_expression <- getMatrixFromProject(proj, useMatrix = "GeneExpressionMatrix")
gene_expr_mat <- assays(gene_expression)[[1]]

print(paste0("The peak matrix has dimensions: ", dim(peak_mat)[[1]], " peaks and ", dim(peak_mat)[[2]], "cells"))
print(paste0("The gene expression matrix contains ", dim(gene_expression) [[1]], " genes"))

```


```{r}
metadata_p2g_atac <- as.data.frame(metadata(p2g)[[1]])
metadata_p2g_rna <- as.data.frame(metadata(p2g)[[2]])


p2g_matrix_from_df <- function(p2g, metadata_p2g_rna, metadata_p2g_atac){
  # add chrosome identifiers to p2g dataframe
  metadata_p2g_atac <- metadata_p2g_atac %>%  rownames_to_column("idxATAC") %>% 
  group_by(seqnames) %>% 
  # create idx per chromosome
  mutate(idx = seq_along(seqnames)) %>% ungroup() %>% 
  # create chr & index combination column, will be needed, since this is how we name peak matrix
  tidyr::unite(chr_idx, seqnames, idx, remove = FALSE, sep = "_")
  
  
  #fitler to get only the peaks which we have in the p2g links dataframe
  atac_chr_idx <- metadata_p2g_atac %>% filter(idxATAC %in% p2g$idxATAC) %>%
    select(idxATAC, chr_idx) 
  
  atac_chr_idx$idxATAC <- as.integer(atac_chr_idx$idxATAC)
  
  # now we can merge this with the p2g dataframe based on the idxATAC column
  p2g_idx <- inner_join(atac_chr_idx, p2g, by = "idxATAC",copy = TRUE)
  
  # add gene names to p2g dataframe
  # get gene names
  metadata_p2g_rna <- as.data.frame(metadata(p2g)[[2]])
  rna_idx <- metadata_p2g_rna %>% rownames_to_column("idxRNA") %>% 
    group_by(seqnames) %>% 
    mutate(idx = seq_along(seqnames)) %>% ungroup() %>% 
    tidyr::unite(chr_idx_rna, seqnames, idx, remove = FALSE, sep = "_") %>% 
    filter(idxRNA %in% p2g_idx$idxRNA) %>% 
    select(idxRNA, name, chr_idx_rna) %>% rename("gene" = name)
  
  rna_idx$idxRNA <- as.integer(rna_idx$idxRNA)
  
  # combine this with the p2g dataframe
  p2g_idx <- inner_join(rna_idx, p2g_idx, by = "idxRNA", copy = TRUE)
  
  # create p2g link matrix
  # create column and row indices for the unique peak and rna indices
  p2g_idx$ColIndex <- match(p2g_idx$chr_idx, unique(p2g_idx$chr_idx))
  p2g_idx$RowIndex <- match(p2g_idx$gene, unique(p2g_idx$gene))
  
  # create a sparse matrix with the correlation values
  p2g_mat <- sparseMatrix(i = p2g_idx$RowIndex, 
                          j = p2g_idx$ColIndex,
                          x = p2g_idx$pos_values,
                          dims = c(length(unique(p2g_idx$gene)),
                                   length(unique(p2g_idx$chr_idx))),
                          dimnames = list(unique(p2g_idx$gene), 
                                          unique(p2g_idx$chr_idx)))
  
  print(paste0("The sparse matrix of correlations has dimensions: ", dim(p2g_mat)[[1]], " genes and ", dim(p2g_mat)[[2]], " peaks."))
  print(paste0("Minimum value: ", min(p2g_mat)))
  print(paste0("The maximum value: ", max(p2g_mat)))
  
  return(p2g_mat)
}
```


```{r}
gene_activity_scores <- function(peak_mat, p2g_mat) {
  peak_mat_subset <- peak_mat[colnames(p2g_mat), ]
  # normalize the p2g matrix by the total number of peaks linked to each gene
  p2g_mat <- p2g_mat / rowSums(p2g_mat)
  print(paste0("normalized the p2g matrix"))
  # Now we can compute a weighted sum of peak2gene correlations for each
  # peak and gene
  scores <- p2g_mat %*% peak_mat_subset
  print(paste0("Computed weightes sum of peaks for each gene and cell"))
  # create a dataframe for computing the linear model
  linear_model_df <- data.frame(cell = colnames(scores),
                               total_activity = colSums(scores),
                               total_sites = colSums(peak_mat_subset))
  # compute a linear model
  activity_model <- stats::lm(log(total_activity) ~ log(total_sites),
                            data = linear_model_df)
  # extract the fitted model
  linear_model_df$fitted_curve <- exp(as.vector(predict(activity_model,
                                                         type = "response")))
  # compute size factors from fitted model
  size_factors <- mean(linear_model_df$fitted_curve) / linear_model_df$fitted_curve
  # create diagonal matrix containing the size factors
  size_factors_mat <- Matrix::Diagonal(x = size_factors)
  row.names(size_factors_mat) <- linear_model_df$cell
  # normalize by library depth size factors
  norm_scores <- Matrix::t(size_factors_mat %*% Matrix::t(scores))
  print(paste0("Normalized for library size"))
  # exponentiate, because RNA counts are log-normally distributed
  norm_scores@x <- pmin(1e9, exp(norm_scores@x) - 1)
  print(paste0("Exponentiated matrix"))
  
  # free some memory
  rm(peak_mat_subset)
  rm(activity_model)
  rm(scores)
  gc(reset = TRUE)

  # scale with total activity scores again
  scale_factors <- Matrix::Diagonal(x = 1/Matrix::colSums(norm_scores))
  print(paste0("Divided by total activity to get value between zero and one"))
  
  final_scores <- Matrix::t(scale_factors %*% Matrix::t(norm_scores))

  return(final_scores)

}

```


The p2g dataframe has a metadata component which contains a Granges object of 
the relevant peaks. The idxATAC column of the p2g dataframe corresponds to this 
Granges object. Below we extract the chr & chr_idx for later. We will also 
extract the gene names based on the idx. 

```{r}
metadata_p2g <- as.data.frame(metadata(p2g)[[1]])
metadata_p2g <- metadata_p2g %>%  rownames_to_column("idxATAC") %>% 
  group_by(seqnames) %>% 
  # create idx per chromosome
  mutate(idx = seq_along(seqnames)) %>% ungroup() %>% 
  # create chr & index combination column, will be needed, since this is how we name peak matrix
  tidyr::unite(chr_idx, seqnames, idx, remove = FALSE, sep = "_")


#fitler to get only the peaks which we have in the p2g links dataframe
atac_chr_idx <- metadata_p2g %>% filter(idxATAC %in% p2g_filt$idxATAC) %>%
  select(idxATAC, chr_idx) 

atac_chr_idx$idxATAC <- as.integer(atac_chr_idx$idxATAC)

# now we can merge this with the p2g dataframe based on the idxATAC column
p2g_idx <- inner_join(atac_chr_idx, p2g_filt, by = "idxATAC",copy = TRUE)
```


```{r}
# get gene names
metadata_p2g_rna <- as.data.frame(metadata(p2g)[[2]])
rna_idx <- metadata_p2g_rna %>% rownames_to_column("idxRNA") %>% 
  group_by(seqnames) %>% 
  mutate(idx = seq_along(seqnames)) %>% ungroup() %>% 
  tidyr::unite(chr_idx_rna, seqnames, idx, remove = FALSE, sep = "_") %>% 
  filter(idxRNA %in% p2g_idx$idxRNA) %>% 
  select(idxRNA, name, chr_idx_rna) %>% rename("gene" = name)

rna_idx$idxRNA <- as.integer(rna_idx$idxRNA)

# combine this with the p2g dataframe
p2g_idx <- inner_join(rna_idx, p2g_idx, by = "idxRNA", copy = TRUE)
```

Create a peak-to-gene link matrix:

```{r}
# create column and row indices for the unique peak and rna indices
p2g_idx$ColIndex <- match(p2g_idx$chr_idx, unique(p2g_idx$chr_idx))
p2g_idx$RowIndex <- match(p2g_idx$gene, unique(p2g_idx$gene))

# create a sparse matrix with the correlation values
p2g_mat <- sparseMatrix(i = p2g_idx$RowIndex, 
                        j = p2g_idx$ColIndex,
                        x = p2g_idx$pos_values,
                        dims = c(length(unique(p2g_idx$gene)),
                                 length(unique(p2g_idx$chr_idx))),
                        dimnames = list(unique(p2g_idx$gene), 
                                        unique(p2g_idx$chr_idx)))

print(paste0("The sparse matrix of correlations has dimensions: ", dim(p2g_mat)[[1]], " genes and ", dim(p2g_mat)[[2]], " peaks."))
print(paste0("Minimum value: ", min(p2g_mat)))
print(paste0("The maximum value: ", max(p2g_mat)))
```

# Gene activity score computation

Now we will calculate a weighted sum of accessible peaks in a cell. The weights
are the corresponding correlation values between a particular peak $j$ and gene
$k$. $A\in\mathbb{R}^{JxI}$ is the binary peak accessibility matrix with $A_{ji} = 1$
if peak $j$ is found in cell $i$ and  $A_{ji} = 0$ otherwise.

$R_{ki} = \sum{U_{kj}A_{ji}}$, where $R_{ki}$ corresponds to the gene activity 
score of gene $k$ in cell $i$.

$R = U I$ with $R \in \mathbb{R}^{KxI}$. 

Below you can see the first few entries of this gene activity score matrix. Note 
that some entries are negative.

Try to simply divide by row sums: $\frac{U_{kj}}{\sum_{j}{U_{kj}}}$.



```{r}
# we subset the accessibility matrix to contain only peaks which are also 
# included in our p2g matrix
#peak_mat_subset <- peak_mat[colnames(p2g_mat), ]


rm(atac_chr_idx)
rm(metadata_p2g)
rm(metadata_p2g_rna)
rm(p2g_filt)
rm(p2g)
rm(p2g_idx)
rm(proj)
rm(peaks)
rm(peakset)
rm(gene_expression)
gc(reset = TRUE)




peak_mat_subset <- peak_mat[colnames(p2g_mat), ]
rm(peak_mat)

# normalize the p2g matrix by the total number of peaks linked to each gene
p2g_mat_norm <- p2g_mat / rowSums(p2g_mat)

data.frame(total_activity = rowSums(peak_mat_subset)) %>%
         ggplot(aes(x = (total_activity))) + geom_histogram(bins = 100) +
  labs(x = "total accessibility across cells")

data.frame(total_activity = rowSums(peak_mat_subset)) %>%
         ggplot(aes(x = log(total_activity))) + geom_histogram(bins = 100) +
  labs(x = "log total accessibility across cells")

data.frame(total_activity = rowSums(gene_expr_mat)) %>%
         ggplot(aes(x = (total_activity))) + geom_histogram(bins = 100) +
  labs(x = "total accessibility across cells")

data.frame(total_activity = rowSums(gene_expr_mat)) %>%
         ggplot(aes(x = log(total_activity))) + geom_histogram(bins = 100) +
  labs(x = "log total accessibility across cells")

# Now we can compute a weighted sum of peak2gene correlations for each
# peak and gene
scores <- p2g_mat_norm %*% peak_mat_subset
```


### Divide by total activity


```#{r}
norm <- Matrix::t(Matrix::t(corr_sum) / 
                    Matrix::colSums(corr_sum))
sc_factor <- 1e4
pseudocount <- 1
#lognorm <- log(norm * sc_factor + pseudocount)
```


### Linear regression to account for differences in library size.

The aggregate correlation values will depend on scATAC-seq library depth in each 
cell. We can capture this relationship with a linear regression. 

$\log{\sum_{k}{r_k} = \beta_0 + \beta_A \log(\sum_j{A_{ji}})}$

```{r}
# create a dataframe for computing the linear model
normalization_df <- data.frame(cell = colnames(scores),
                               total_activity = colSums(scores),
                               total_sites = colSums(peak_mat_subset))

#plot the relationship between library depth and activity scores
normalization_df %>% ggplot(aes(x = total_sites, y = total_activity)) + geom_point() +
  labs(title = "Relationship between activity scores and library depth", 
       y = "total activity", x = "peaks")



# compute a linear model
activity_model <- stats::lm(log(total_activity) ~ log(total_sites),
                            data = normalization_df)


# extract the fitted model
normalization_df$fitted_curve <- exp(as.vector(predict(activity_model, type = "response")))

# plot the fitted model
normalization_df %>% ggplot(aes(x = total_sites, y = fitted_curve)) + geom_point() +
  labs(title = "Fitted Model capturing relationship between activity scores and library depth", y = "Fitted curve - activity scores", x = "number of peaks") +
  geom_hline(yintercept = mean(normalization_df$fitted_curve), color = "orange")

#rm(normalization_df)
gc(reset = TRUE)
```


### SizeFactors

The output of the linear model is used to scale the accessibility we computed
above.

$\tilde{R}_{ki}= R_{ki} * \frac{\sum_{i}{r_i}}{r_i}$

If an activity score $R_{ki}$ for a given cell $i$ is smaller than the average activity score, the 
size factor will be >1 and if the activity score  is bigger than the average 
activity score the size factor will be <1. After correcting with the size factors we would expect the linear increase in activity with increasing number of accessibile 
sites to be corrected for. 

In the plot below we can see that we have corrected for this linear relationship.

```{r}
# each cell is scaled by a factor corresponding to its library depth
#size_factors <- log(normalization_df$fitted_curve) / mean(log(normalization_df$fitted_curve))
#size_factors <- log(sum(normalization_df$fitted_curve)) / log(normalization_df$fitted_curve) 

size_factors <- mean(normalization_df$fitted_curve) / normalization_df$fitted_curve


ggplot(as.data.frame(size_factors), aes(x = size_factors)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = 1, color="orange", alpha=0.8)

#size_factors_mat <- Matrix::Diagonal(x = 1/size_factors)
size_factors_mat <- Matrix::Diagonal(x = size_factors)

row.names(size_factors_mat) <- normalization_df$cell
size_factors_mat[1:5, 1:5]
# adjust scores by size factors 
new_scores <- Matrix::t(size_factors_mat %*% Matrix::t(scores))
print(new_scores[1:5, 1:5])

# plot the total sites vs total activities after correction
df <- data.frame(cell = colnames(new_scores),
                               total_activity = colSums(new_scores),
                               total_sites = colSums(peak_mat_subset))
df %>% ggplot(aes(x = total_sites, y = total_activity)) + 
  geom_point(size=0.2, alpha=0.4) +
  ggside::geom_ysidedensity() +
  geom_hline(yintercept = mean(df$total_activity), color = "orange")

print(paste0("Mean activity after correction (expected to be the same as the mean of the fitted activity: ", mean(df$total_activity)))
```



### Transform

Then we exponentiate the values, because the RNA-seq measures are typically 
log-normally distributed. To create relative exponentiated gene accessibility
values, we scale them by the total exponentiated gene accesibility values. We 
get our gene activity scores $G_{ki}$.

Because for log transformation of gene expression counts we add a pseudocount, 
we need to subtract one here:

$e^{\tilde{R}} - 1$

$G_{ki} = \frac{e^{\tilde{R}_{ki}}-1}{\sum_k{e^{\tilde{R}_{ki}}-1}}$

Below you can see the first few entries of the exponentiated values, before scaling.

```{r}
# plot the total sites vs total activities after correction
# this will set an upper bound to 1e9
new_scores@x <- pmin(1e9, exp(new_scores@x) - 1)
print(new_scores[1:5, 1:5])
```



```{r}
min(new_scores)
max(new_scores)
```


```{r}
rm(gene_expr_mat)
rm(rna_idx)
rm(normalization_df)
gc(reset = TRUE)
```


### Scaling by total gene activity in a given cell

Here, you can see the first few entries of the transformed matrix. The column
sum is now 1. 

```{r}
# scale with total activity scores again
scale_factors <- Matrix::Diagonal(x = 1/Matrix::colSums(new_scores))

new_scores <- Matrix::t(scale_factors %*% Matrix::t(new_scores))
print(new_scores[1:5, 1:5])

# Check whether the column sum is 1:
colSums(new_scores)%>% head


# plot the total sites vs total activities after correction
corrected_df <- data.frame(cell = colnames(new_scores),
                               total_activity = colSums(new_scores),
                               total_sites = colSums(peak_mat_subset))
corrected_df %>% ggplot(aes(x = total_sites, y = total_activity)) + geom_point()


print(paste0("The minimum value is: ", min(new_scores)))
print(paste0("The maximum value is: ", max(new_scores)))
```

```{r}
rm(p2g_mat)
rm(peak_mat_subset)
rm(scores)
gc(reset = TRUE)
```




```#{r}
tibble::tibble(
  gene = rownames(new_scores),
  mean = Matrix::rowMeans(new_scores), 
  detection_rate = Matrix::rowMeans(new_scores > 0), 
  var = MatrixGenerics::rowVars(new_scores)) %>%
  mutate(log10_mean = log10(mean),
         log10_var = log10(var)) %>% 
  ggplot(aes(log10_mean, log10_var)) + 
  geom_point(alpha = 0.3, shape = 16) + 
  geom_density_2d(size = 0.3, bins = 6) + 
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Log10 Mean",y = "Log10 Var")
```







# Add gene score matrix to ArchR project


```#{r}
proj <- loadArchRProject("11_added_Ricards_peaks/")

# get the Granges object from the Archr project
archr_granges <-getGenes(proj)

# filter for genes which are included in the gene expression matrix
granges <- subset(archr_granges, symbol %in% rownames(new_scores))
```




We will now add the gene expression matrix to the ArchR object. First, we have to 
convert the Single Cell Experiment to a Summarized Experiment.

```#{r}
# create summarized experiment of the scRNA-seq
seRNA <- SummarizedExperiment(assays = list(counts=new_scores),
                          colData = colnames(new_scores),
                          rowRanges = granges)
seRNA


# add RNA gene expression matrix to the ArchR object
# we need to set scaleTo to one, since it is 1000 by default!
proj <- addGeneExpressionMatrix(input = proj, seRNA = seRNA, force = TRUE, 
                                scaleTo = 1)


# save the ArchR object with the added gene expression matrix
#saveArchRProject(proj, outputDirectory = "13_added_gene_score_from_p2g/", load = FALSE)

#saveArchRProject(proj, outputDirectory = "15_gene_scores_from_p2g_as_gene_expr_matrix")
```


```{r}
proj <- loadArchRProject("15_gene_scores_from_p2g_as_gene_expr_matrix/")

marker_genes <- c("Lamb1",  "Sparc", "Elf5", "Ascl2", "Tfap2c", "Ttr",
                  "Apoa2", "Apoe", "Cystm1", "Emb", "Spink1",  "Krt19",
                  "Dkk1", "Grhl3", "Trp63", "Grhl2",  "Pax6", "Pax2",
                  "En1", "Foxd3", "Tfap2a", "Pax3", "Sox9",
                  "Six3", "Hesx1", "Irx3", "Sox2", "Hoxb9", "Cdx4", 
                  "Hes3", "Hba-a2", "Hba-a1",  "Hbb-bh1", "Gata1", "Cited4",
                   "Cdh5", "Pecam1", "Anxa5", "Etv2", "Igf2",
                  "Krt8", "Krt18", "Pmp22", "Ahnak", "Bmp4", "Tbx4", "Hoxa11", 
                  "Hoxa10", "Tnnt2", "Myl4",  "Myl7", "Acta2", 
                  "Smarcd3", "Tcf21", "Tbx6", "Dll1", "Aldh1a2", "Tcf15", 
                  "Meox1", "Tbx1", "Gbx2", "Cdx1", "Hoxb1", "Hes7", "Osr1", 
                  "Mesp2", "Lefty2", "Mesp1", "Cer1",  "Chrd", "T", 
                  "Foxa2", "Pax7", "Fgf8", "Lhx1", "Gsc", "Mixl1", "Otx2", "Hhex",
                   "Ifitm3", "Nkx1-2", "Eomes", "Nanog", "Utf1", 
                  "Epcam", "Pou5f1")
```

### Gene Scores UMAPs

```{r, fig.width=10, fig.height=50}

#marker_genes <- c("Hba-a2", "Hba-a1",  "Hbb-bh1")
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneExpressionMatrix", 
    name = marker_genes, 
    embedding = "UMAP",
    quantCut = c(0.01, 0.95),
    imputeWeights = getImputeWeights(proj)
)


plots <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})

do.call(cowplot::plot_grid, c(list(ncol = 3),plots))

```

```#{r}
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneExpressionMatrix", 
    name = "Gata6", 
    embedding = "UMAP",
    quantCut = c(0.01, 0.95)
)

p
```

### Gene Expression Matrix UMAPs
```{r}
rm(proj)
rm(new_scores)
gc(reset = TRUE)
proj <- loadArchRProject("12_Ricards_peaks_ChromVar/")
scores <- getMatrixFromProject(proj, useMatrix = "GeneExpressionMatrix")
new_scores <- assays(scores)[[1]]
rm(scores)
gc(reset = TRUE)

#marker_genes <- c("Hba-a2", "Hba-a1",  "Hbb-bh1")
p <- plotEmbedding(
    ArchRProj = proj, 
    colorBy = "GeneExpressionMatrix", 
    name = marker_genes, 
    embedding = "UMAP",
    quantCut = c(0.01, 0.95),
    imputeWeights = getImputeWeights(proj)
)


plots <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})

do.call(cowplot::plot_grid, c(list(ncol = 3),plots))


```



# Correlation between gene activity score and gene expression:

## Create cell aggregates

```{r}
gene_expression <- getMatrixFromProject(proj, useMatrix = "GeneExpressionMatrix")
# extract gene expression matrix from gene expression object
gene_expr_mat <- assays(gene_expression)[[1]]
rownames(gene_expr_mat) <- rowData(gene_expression)$name
rm(gene_expression)
gc(reset = TRUE)

# randomly sample 5% of cells
sample_frac <- .05

# out of all cells in the gene expression matrix randomly sample 5% without replacement
sampled_cells <- sample(1:ncol(gene_expr_mat), sample_frac * ncol(gene_expr_mat), replace = FALSE)
print(paste0("Sampling ", length(sampled_cells), " cells."))
```



## Add PCA




```{r}
library(Seurat)
#atac_seurat <- readRDS("atac_Seurat_object")
rna_seurat <- readRDS("Seurat_objects/rna_Seurat_object")

# get the cell embeddings
cell_embeddings <- Embeddings(object = rna_seurat, reduction = "pca")#[1:5, 1:5]


# get gene score matrix
#proj <- loadArchRProject("12_Ricards_peaks_ChromVar/")
#gene_scores <- getMatrixFromProject(proj, useMatrix = "GeneScoreMatrix")
```


First, we want to use only genes which we find in our newly computed gene
activity scores from p2g links, the gene expression matrix and the imputed gene
scores from scATAC.

```{r}

#scores <- read.table("gene_scores_from_p2g/norm_simple_correction_no_scaling")
```


```{r}
gc(reset = TRUE)
# create nearest neighbor graph
k <- 50

# use only cells we also have in gene expr/score matrix and the first 30 PCs
nn <- RANN::nn2(cell_embeddings[colnames(gene_expr_mat), 1:30], k = k)

# get genes of interest, the top 2000 most highly variable features
hvg <- rna_seurat@assays$originalexp@var.features

# we are left with 1772 genes
hvg <- hvg[hvg %in% rownames(new_scores)]
length(hvg)


# initialize matrix to store aggregate values
rna_score <- matrix(0, nrow = length(hvg), ncol = length(sampled_cells))
#atac_score <- matrix(0, nrow = length(hvg), ncol = length(sampled_cells))
p2g_score <- matrix(0, nrow = length(hvg), ncol = length(sampled_cells))
rownames(rna_score) <- hvg
#rownames(atac_score) <- hvg
rownames(p2g_score) <- hvg

# reorder columns to match gene expression matrix columns
#reord_gene_scores <- gene_score_mat[, colnames(gene_expr_mat)]
reord_p2g_scores <- new_scores[, colnames(gene_expr_mat)]
# 
# 
for(i in 1:length(sampled_cells)) {
# check progress
if (i %% 100 == 0) print(i)

# get current cell
cell <- sampled_cells[i]

# aggregate the rna score for one gene and all nearest neighbors of the first
# cell in our aggregate group of cells
rna_score[hvg, i] <- sparseMatrixStats::rowMeans2(
  gene_expr_mat[hvg, nn$nn.idx[cell, ]]
)

  # aggregate the atac score
p2g_score[hvg,i] <- sparseMatrixStats::rowMeans2(
  reord_p2g_scores[hvg, nn$nn.idx[cell,]]
  )
}

# write.table(rna_score, "gene_scores_from_p2g/wrong_computation_good/rna_score")
# write.table(p2g_score, "gene_scores_from_p2g/wrong_computation_good/p2g_score")

```


```{r}
rm(rna_seurat)
rm(gene_expr_mat)
gc(reset= TRUE)
```


```{r}
hvg <- hvg[hvg %in%rownames(rna_score)]

comp_stats <- rbind(rna_score %>% set_rownames(paste0(hvg, "_#_rna_score")),
                    p2g_score %>% set_rownames(paste0(hvg, "_#_p2g_score"))) %>% 
  t() %>% 
  as.matrix() %>% 
  tibble::as_tibble() %>% 
  mutate(cell_name = sampled_cells) %>% 
  pivot_longer(cols = !cell_name, names_to = "name", values_to = "vals") %>%
  separate(name, sep = "_#_", into= c("gene", "statistic")) %>% 
  pivot_wider(names_from = statistic, values_from = vals)
```


```{r, fig.widht = 5, fig.height = 20}
marker_genes <- c("Lamb1",  "Sparc", "Elf5", "Ascl2", "Tfap2c", "Ttr",
                  "Apoa2", "Apoe", "Cystm1", "Emb", "Spink1",  "Krt19",
                  "Dkk1", "Grhl3", "Trp63", "Grhl2",  "Pax6", "Pax2",
                  "En1", "Foxd3", "Tfap2a", "Pax3", "Sox9",
                  "Six3", "Hesx1", "Irx3", "Sox2", "Hoxb9", "Cdx4", 
                  "Hes3", "Hba-a2", "Hba-a1",  "Hbb-bh1", "Gata1", "Cited4",
                   "Cdh5", "Pecam1", "Anxa5", "Etv2", "Igf2",
                  "Krt8", "Krt18", "Pmp22", "Ahnak", "Bmp4", "Tbx4", "Hoxa11", 
                  "Hoxa10", "Tnnt2", "Myl4",  "Myl7", "Acta2", 
                  "Smarcd3", "Tcf21", "Tbx6", "Dll1", "Aldh1a2", "Tcf15", 
                  "Meox1", "Tbx1", "Gbx2", "Cdx1", "Hoxb1", "Hes7", "Osr1", 
                  "Mesp2", "Lefty2", "Mesp1", "Cer1",  "Chrd", "T", 
                  "Foxa2", "Pax7", "Fgf8", "Lhx1", "Gsc", "Mixl1", "Otx2", "Hhex",
                   "Ifitm3", "Nkx1-2", "Eomes", "Nanog", "Utf1", 
                  "Epcam", "Pou5f1")


comp_stats %>%
  filter(gene %in% marker_genes) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free", ncol = 5) +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```


```{r}
#n <- 10
#genes_oi <- hvg[1:n]
genes_oi <- c("Hba-a2", "Hba-a1",  "Hbb-bh1", "Gata1", "Cited4",  "Foxa2",
              "Cdh5", "Pecam1", "Anxa5", "Etv2")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```


                  
```{r}
genes_oi <- c("Lamb1",  "Sparc", "Elf5", "Ascl2", "Tfap2c", "Ttr",
                  "Apoa2", "Apoe", "Cystm1", "Emb")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```
                  




```{r}
genes_oi <- c("Spink1",  "Krt19",
                  "Dkk1", "Grhl3", "Trp63", "Grhl2",  "Pax6", "Pax2",
                  "En1", "Foxd3")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```

```{r}
genes_oi <- c( "Tfap2a", "Pax3", "Sox9",
                  "Six3", "Hesx1", "Irx3", "Sox2", "Hoxb9", "Cdx4", 
                  "Hes3")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```






```{r}
genes_oi <- c("Igf2",
                  "Krt8", "Krt18", "Pmp22", "Ahnak", "Bmp4", "Tbx4", "Hoxa11", 
                  "Hoxa10", "Tnnt2")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```

```{r}
genes_oi <- c("Myl4",  "Myl7", "Acta2", 
                  "Smarcd3", "Tcf21", "Tbx6", "Dll1", "Aldh1a2", "Tcf15", 
                  "Meox1")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```


```{r}
genes_oi <- c("Tbx1", "Gbx2", "Cdx1", "Hoxb1", "Hes7", "Osr1", 
                  "Mesp2", "Lefty2", "Mesp1", "Cer1")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```


```{r}
genes_oi <- c("Chrd", "T", 
                  "Foxa2", "Pax7", "Fgf8", "Lhx1", "Gsc", "Mixl1", "Otx2", "Hhex")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```



```{r}
genes_oi <- c("Ifitm3", "Nkx1-2", "Eomes", "Nanog", "Utf1", 
                  "Epcam", "Pou5f1")
comp_stats %>%
  filter(gene %in% genes_oi) %>%
  ggplot(aes(x = rna_score, y = p2g_score)) +
  geom_point(size=0.4, alpha=0.4) +
  geom_smooth(formula = y ~ x, method = "lm", size=0.1) +
  #ggpubr::stat_cor(aes(label = ..r.label..), method = "pearson", r.digits=2) +
  facet_wrap(~ gene, scales="free") +
  labs(x = "RNA Counts", y = "P2G Gene Activity Score")
```
```#{r}
getCellColData(proj) %>% as.data.frame() %>% 
  mutate("Hba_a2" = new_scores["Hba-a2", ]) %>%
  ggplot() +
  geom_boxplot(aes(x = Hba_a2, y = celltypes))# %>% 
  
```


```#{r}
getCellColData(proj) %>% as.data.frame()%>% ggplot() +
  geom_bar(aes(x = celltypes, fill = celltypes)) +
  scale_fill_manual(values = col) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none")
```

```#{r}
colPalette_celltypes = c('#532C8A',
 '#c19f70',
 '#f9decf',
 '#c9a997',
 '#B51D8D',
 '#3F84AA',
 '#9e6762',
 '#354E23',
 '#F397C0',
 '#ff891c',
 '#635547',
 '#C72228',
 '#f79083',
 '#EF4E22',
 '#989898',
 '#7F6874',
 '#8870ad',
 '#647a4f',
 '#EF5A9D',
 '#FBBE92',
 '#139992',
 '#cc7818',
 '#DFCDE4',
 '#8EC792',
 '#C594BF',
 '#C3C388',
 '#0F4A9C',
 '#FACB12',
 '#8DB5CE',
 '#1A1A1A',
 '#C9EBFB',
 '#DABE99',
 '#65A83E',
 '#005579',
 '#CDE088',
 '#f7f79e',
 '#F6BFCB')

celltypes <- (as.data.frame(getCellColData(proj)) %>% group_by(celltypes) %>% 
 summarise(n = n()))$celltypes

col <- setNames(colPalette_celltypes, celltypes)
```

```#{r}
getCellColData(proj) %>% as.data.frame() %>% 
  mutate("Hba_a2" = new_scores["Hba-a2", ]) %>%
  ggplot() +
  geom_bar(aes(x = celltypes, y = Hba_a2, fill = celltypes), stat = "identity") +
  scale_fill_manual(values = col) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none")


getCellColData(proj) %>% as.data.frame() %>% 
  mutate("Hba_a2" = new_scores["Hba-a2", ]) %>%
  group_by(celltypes) %>% 
  mutate(counts = n()) %>% ungroup %>% 
  mutate("scaled_Hba_a2" = Hba_a2 / counts) %>% 
  ggplot() +
  geom_boxplot(aes(x = celltypes, y = log10(scaled_Hba_a2))) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

  
  geom_bar(aes(x = celltypes, y = scaled_Hba_a2, fill = celltypes), stat = "identity") +
  scale_fill_manual(values = col) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none")
```



