---
title: "test"
output: html_document
---

Find row-wise distance between two GRanges objects:

```{r}
query <- GRanges("A", IRanges(c(1,2,10, 13), c(5,8,11, 15)))
query
subject <- GRanges("A", IRanges(c(6, 5, 13, 10), c(10, 10, 15, 11))) 
subject
GenomicRanges::distance(query, subject)
```


Find overlapping ranges

```{r}
query <- IRanges(c(1, 4, 9), c(5, 7, 10))
query
subject <- IRanges(c(2, 2, 10), c(2, 3, 12))
subject

overlaps <- findOverlaps(query, subject)
overlaps
```

```{r}

findOverlaps(query, subject, select="first")
findOverlaps(query, subject, select="last")
findOverlaps(query, subject, select="arbitrary")
```


From ArchR Gene activity scores:

The function takes as input a charcter containing the formula/expression.
In the code, they use `parse()` which returns an object of type `expression`  

```{r}

x <- c(2,5,1,3,7,8, 4,3)
 
eval(parse(text="exp(-abs(x)/5000) + exp(-1)"))

# does not work:
#eval("exp(-abs(x)/5000) + exp(-1)"))


parse(text="exp(-abs(x)/5000) + exp(-1)")
```



# Expressions

We want to distinguish between the action of multiplying x by 10 and the results
"40". 

```{r}
x <- 4
y <- x * 10
y
```

The action can be captured using `quote()`. `quote()` returns an *expression*. An
expression is an object which reprsents an action that can be performed in R. Expression
returns a list of expressions which can be extracted as a real expression using 
`parse()`. With `parse()` a string can be converted to an expression. This is primarily
for parsing files of code to disk. Code in a character vector as in the example above,
requires an additional argument, `parse(text = "")`. 

```{r}
z <- quote(y <- x*10)
z

deparse(z)
parse(text = deparse(z))

```

# Find Overlaps

```{r}

gr <- GRanges(
        seqnames=Rle(c("chr1", "chr2", "chr1", "chr3"), c(1, 3, 2, 4)),
        ranges=IRanges(1:10, width=10:1, names=head(letters,10)),
        strand=Rle(strand(c("-", "+", "*", "+", "-")), c(1, 2, 2, 3, 2)),
        score=1:10,
        GC=seq(1, 0, length=10)
      )
gr

gr1 <- GRanges(seqnames="chr2", ranges=IRanges(4:3, 6),
               strand="+", score=5:4, GC=0.45)
gr2 <- GRanges(seqnames=c("chr1", "chr1"),
               ranges=IRanges(c(7,13), width=3),
               strand=c("+", "-"), score=3:4, GC=c(0.3, 0.5))
gr3 <- GRanges(seqnames=c("chr1", "chr2"),
               ranges=IRanges(c(1, 4), c(3, 9)),
               strand=c("-", "-"), score=c(6L, 2L), GC=c(0.4, 0.1))
grl <- GRangesList("gr1"=gr1, "gr2"=gr2, "gr3"=gr3)

gr1

```

```{r}
countOverlaps(gr, gr1)
```


```{r}
findOverlaps(gr, gr1)
```



# Nearest

Whenever multiple ranges overlap, an arbitrary range is chosen

```{r}
query <- GRanges("A", IRanges(5, 15))
query
subject <- GRanges("A", IRanges(c(1, 15), c(5, 19)))
subject
nearest(query, subject)
```

